### 2015-06-06 chapters/pitfalls.tex
adding IPA-pitfall about non-decomposition
reformatted pitfall file formats

Although in most cases canonical equivalence will take care of alternative
encodings of the same character, there are a some cases in which the Unicode
Standard decided against equivalence. This leads to identical characters that
are not equivalent, like <ø> \textsc{latin small letter o with stroke} at
\uni{00F8} and <o̷> a combination of \textsc{latin small letter o} at \uni{006F}
with \textsc{combining short solidus overlay} at \uni{0037}.

The general rule followed is that extensions of Latin characters that are
connected to the character are not separated as combining diacritics. For
example, characters like <ƞ ɲ ɳ> or <ɰ ɱ> are obviously derived from <n> and <m>
respectively, but they are treated like new separate characters in the Unicode
Standard. Likewise, characters like <ø> and <ƈ> are not separated into a base 
character <o> and <c> with an attached combining diacritic.

@@

Unicode is a character encoding standard, but these characters of course
actually appear inside some kind of file. The most basic Unicode-based file
format is pure line-based text, i.e.~strings of Unicode-encoded characters
separated by line breaks (note that these line breaks are what for most people
intuitively corresponds to paragraph breaks). Unfortunately, even within this
apparently basic setting there exist a multitude of variants. In general, these
different possibilities are well-understood in the software industry, and
nowadays they normally do not lead to any problems for the end user. However,
there are some situations in which a user is suddenly confronted with cryptic
questions in the user interface involving abbreviations like \textsc{lf},
\textsc{cr}, \textsc{be}, \textsc{le} or \textsc{bom}. Most prominently this
occurs with exporting or importing data in several software applications from
Microsoft. Basically, there are two different issues involved. First, the
encoding of line breaks and, second, the encoding of the Unicode characters into
\textsc{code units} and the related issue of \textsc{endianness}.

The issue with \textsc{line breaks} originated with the instructions necessary
to direct a printing head to a new line. This involves two movements, known as
\textsc{carriage return (cr)} (returning the printing head to the start of the
line on the page) and \textsc{line feed (lf)} (moving the printing head to the
next line on the page). Physically, these are two different events, but
conceptually together they form one action. In the history of computing, various
encodings of line breaks have been used (e.g.~\textsc{cr+lf}, \textsc{lf+cr},
only \textsc{lf}, or only \textsc{cr}). Currently, all Unix and Unix-derived
systems use only \textsc{lf} as code for a line break, while software from
Microsoft still uses a combination of \textsc{cr} and \textsc{lf}. Today, most
software applications recognize both options, and are able to deal with either
encoding of line breaks (until rather recently this was not the case, and using
the wrong line breaks could lead to unexpected errors). Our impression is that
there is a strong tendency in software development to standardize on the simpler
\textit{only \textsc{lf}} encoding for line breaks, and we suggest that
everybody use this encoding whenever possible.

A code unit is a sequence of bits used to encode character in a encoding. The
issue with code units stems from the question how to separate a stream of binary
ones and zero, i.e.~bits, into chunks representing Unicode characters. Depending
on different use cases, the Unicode Standard offers three different approaches,
called UTF-32, UTF-16 and UTF-8.\footnote{The letters UTF stand for either
\textsc{Unicode Transformation Format} or \textsc{Universal Coded Character Set
Transformation Format}, but the notion of \textit{transformation} is a legacy
notion that does not have meaning anymore. Nevertheless, the designation
\textsc{UTF} (in capitals) has become an official standard designation, but
should probably best be read as simply \textit{Unicode Format}.} The details of
this issue is extensively explained in section 2.5 of the Unicode core
specification. Basically, \textsc{UTF-32} encodes each character in 32 bits (32
\textit{bi}nary un\textit{ts}, i.e.~32 zeros or ones) and is the most
disk-space-consuming variant of the three. However, it is the most efficient
encoding processing-wise, because the computer simply has to separate each
character after 32 bits. In contrast, \textsc{UTF-16} uses only 16 bits per
character, which is sufficient for the large majority of Unicode characters, but
not for all of them. A special system of `surrogates' is defined within the
Unicode Standard to deal with these additional characters. The effect is a more
disk-space efficient encoding (approximately half the size), while adding a
limited computational overhead. Finally, \textsc{UTF-8} is a more complex system
that dynamically encodes each character with the minimally necessary number of
bits, choosing either 8, 16 or 32 bits depending on the character. This
represents again a strong reduction in space (particularly due to the high
frequency of data using ASCII characters that need only 8 bits) at the expense
of even more computation necessary to process such strings. However, because of
the ever growing computational power of modern machines, the processing overhead
is in most practical situations a non-issue, while saving on space is still
useful, particularly for sending texts over the Internet. As an effect, UTF-8
has become the dominant encoding on the World Wide Web. We suggest that
everybody uses UTF-8 as their default encoding.

A related problem is a general issue about how to store information in computer
memory, which is known as \textsc{endianness}. The details of this issue go
beyond the scope of this paper. It suffices to realize that there is a
difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
(LE) storage. The Unicode Standard offers a possibility to explicitly indicate
what kind of storage is used by starting a file with a so-called `Byte Order
Mark' (BOM). However, the Unicode Standard does not require the usage of BOM,
preferring other non-Unicode methods to signal to computers which kind of
endianness is used. This issue only arises with UTF-32 and UTF-16 encodings, and
when using the preferred UTF-8, using a BOM is theoretically possible, but
strongly dispreferred according to the Unicode Standard. We suggest that
everyone tries to prevent the inclusion of BOM in your data.

Together with the topic of normalization (NFC vs.~NFD, see Pitfall 5 above),
this leads to the following nicely cryptic file-encoding recommendation, namely
to preferably use \textsc{UTF-8 in NFC without BOM with LF line breaks},
whenever possible.

>>>>

Although in most cases canonical equivalence will take care of alternative
encodings of the same character, there are a some cases in which the Unicode
Standard decided against equivalence. This leads to identical characters that
are not equivalent, like <ø> \textsc{latin small letter o with stroke} at
\uni{00F8} and <o̷> a combination of \textsc{latin small letter o} at \uni{006F}
with \textsc{combining short solidus overlay} at \uni{0037}.
The general rule followed is that extensions of Latin characters that are
connected to the character are not separated as combining diacritics. For
example, characters like <ƞ ɲ ɳ> or <ɰ ɱ> are obviously derived from <n> and <m>
respectively, but they are treated like new separate characters in the Unicode
Standard. Likewise, characters like <ø> and <ƈ> are not separated into a base 
character <o> and <c> with an attached combining diacritic.

@@

Unicode is a character encoding standard, but these characters of course
actually appear inside some kind of file. The most basic Unicode-based file
format is pure line-based text, i.e.~strings of Unicode-encoded characters
separated by line breaks (note that these line breaks are what for most people
intuitively corresponds to paragraph breaks). Unfortunately, even within this
apparently basic setting there exist a multitude of variants. In general, these
different possibilities are well-understood in the software industry, and
nowadays they normally do not lead to any problems for the end user. However,
there are some situations in which a user is suddenly confronted with cryptic
questions in the user interface involving abbreviations like LF, CR, BE, LE or
BOM. Most prominently this occurs with exporting or importing data in several
software applications from Microsoft. Basically, there are two different issues
involved. First, the encoding of line breaks and, second, the encoding of the
Unicode characters into \textsc{code units} and the related issue of
\textsc{endianness}.

\subsubsection*{Line breaks}

The issue with \textsc{line breaks} originated with the instructions necessary
to direct a printing head to a new line. This involves two movements, known as
\textsc{carriage return} (CR, returning the printing head to the start of the
line on the page) and \textsc{line feed} (LF, moving the printing head to the
next line on the page). Physically, these are two different events, but
conceptually together they form one action. In the history of computing, various
encodings of line breaks have been used (e.g.~CR+LF, LF+CR, only LF, or only CR.
Currently, all Unix and Unix-derived systems use only LF as code for a
line break, while software from Microsoft still uses a combination of
CR and LF. Today, most software applications recognize both
options, and are able to deal with either encoding of line breaks (until rather
recently this was not the case, and using the wrong line breaks could lead to
unexpected errors). Our impression is that there is a strong tendency in
software development to standardize on the simpler \textsc{only LF}
encoding for line breaks, and we suggest that everybody use this encoding
whenever possible.

\subsubsection*{Code units}

The issue with \textsc{code units} stems from the question how to separate a
stream of binary ones and zero, i.e.~bits, into chunks representing Unicode
characters. A code unit is the sequence of bits used to encode a single
character in an encoding. Depending on different use cases, the Unicode Standard
offers three different approaches, called UTF-32, UTF-16 and UTF-8.\footnote{The
letters UTF stand for either \textsc{Unicode Transformation Format} or
\textsc{Universal Coded Character Set Transformation Format}, but the notion of
\textit{transformation} is a legacy notion that does not have meaning anymore.
Nevertheless, the designation \textsc{UTF} (in capitals) has become an official
standard designation, but should probably best be read as simply Unicode
Format.} The details of this issue is extensively explained in section 2.5 of
the Unicode Core Specification \cite{Unicode2014}.

Basically, \textsc{UTF-32} encodes each character in 32 bits (32 \textit{bi}nary
uni\textit{ts}, i.e.~32 zeros or ones) and is the most disk-space-consuming
variant of the three. However, it is the most efficient encoding
processing-wise, because the computer simply has to separate each character
after 32 bits.

In contrast, \textsc{UTF-16} uses only 16 bits per character, which is
sufficient for the large majority of Unicode characters, but not for all of
them. A special system of \textsc{surrogates} is defined within the Unicode Standard to
deal with these additional characters. The effect is a more disk-space efficient
encoding (approximately half the size), while adding a limited computational
overhead to manage the surrogates.

Finally, \textsc{UTF-8} is a more complex system that dynamically
encodes each character with the minimally necessary number of bits, choosing
either 8, 16 or 32 bits depending on the character. This represents again a
strong reduction in space (particularly due to the high frequency of data using
ASCII characters, which need only 8 bits) at the expense of even more computation
necessary to process such strings. However, because of the ever growing
computational power of modern machines, the processing overhead is in most
practical situations a non-issue, while saving on space is still useful,
particularly for sending texts over the Internet. As an effect, UTF-8 has become
the dominant encoding on the World Wide Web. We suggest that everybody uses
UTF-8 as their default encoding.

A related problem is a general issue about how to store information in computer
memory, which is known as \textsc{endianness}. The details of this issue go
beyond the scope of this book. It suffices to realize that there is a
difference between \textsc{big-endian} (BE) storage and \textsc{little-endian}
(LE) storage. The Unicode Standard offers a possibility to explicitly indicate
what kind of storage is used by starting a file with a so-called \textsc{Byte Order
Mark} (BOM). However, the Unicode Standard does not require the usage of BOM,
preferring other non-Unicode methods to signal to computers which kind of
endianness is used. This issue only arises with UTF-32 and UTF-16 encodings.
When using the preferred UTF-8, using a BOM is theoretically possible, but
strongly dispreferred according to the Unicode Standard. We suggest that
everyone tries to prevent the inclusion of BOM in your data.

\

\noindent Together with the topic of normalization (NFC vs.~NFD, see 
Section~\ref{pitfall-canonical-equivalence} above),
this leads to the following nicely cryptic file-encoding recommendation, namely
to preferably use UTF-8 in NFC without BOM with LF line breaks,
whenever possible.
